\replaced[id=siyu]{Separating random and deterministic sources of computational noise in explore-exploit decisions}{The nature of decision noise in random exploration}

### 2023-12-21 manuscript/Older drafts/2023 unused - overleaf/manuscript_DeterministicRandomNoise.tex
u

% control the fontsize
\PassOptionsToPackage{textsize=scriptsize}{todonotes}
%\usepackage[final]{changes}
\usepackage{changes}
\definechangesauthor[name={Siyu Wang}, color=red]{siyu}

@@

\title{\replaced[id=siyu]{Separating random and deterministic sources of computational noises in explore-exploit decisions}{The nature of decision noise in random exploration}}
\author[1,\textcurrency]{Siyu Wang}
\author[1,2,3]{Robert C. Wilson}

@@

	\newpage
	\begin{abstract}
		Human decision making is inherently variable. While this variability is often seen as a sign of suboptimality behavior, recent work suggests that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. A little randomness in these `explore-exploit' decisions is remarkably effective as it encourages us to explore options we might otherwise ignore. Recent work suggests that people may actually use such `random exploration' in practice, increasing their behavioral variability when it is more valuable to explore. 
		%Other work, however, suggests that variability in human behavior is not always random, and is instead driven by deterministic factors (such as repeating past actions) 
		A key question is whether the variability in random exploration is actually random. That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. \added[id=siyu]{We built a novel computational model which could detect and separate previously unobserved deterministic processes from random noise, without having to make a priori assumptions about the nature of the deterministic process. } In particular, we find evidence that around 15$\%$ of the variability in random exploration can be accounted for by deterministic processing of the stimulus. This still leaves open the possibility that much of random exploration is truly `random', but narrows the window of opportunity for \added[id=siyu]{true} stochastic processes to explain this behavior. \added[id=siyu]{Moreover, our results suggest that both deterministic noise and random noise change proportionally to each other as the planning horizon changes. This sheds light on a common noise gating mechanism that is at play in random exploration. Our work takes us one step closer to understanding of the nature of decision noise in explore-exploit decisions.}

@@

	Imagine trying to decide where to go to dinner on a date, you can go to your favorite restaurant, the one you both really enjoy and always go to, or you can try a new restaurant that you know nothing about. Such decisions, in which we must choose between a well-known `exploit' option and a lesser known `explore' option, are known as explore-exploit decisions.  From a theoretical perspective, making optimal explore-exploit choices, i.e. choices that maximize long-term reward, is computationally intractable in most cases \citep{eegittins74, eeBasu18}. In part because of this difficulty, there is considerable interest in how humans and animals solve the explore-exploit dilemma in practice \added[id=siyu]{\citep{Mehlhorn15, SCHULZ20197,WILSON202149}}.
	%\citep{eeauer02,eegittins79,eekrebs78,eethompson33, eewatkins89, eebridle90,eemeyer95, eebanks97,eefrank09, eesteyvers09, eelee11, eepl12,eezhang13,eedaw06, eepl11, wilson2014}

	One particularly effective strategy for solving the explore-exploit dilemma is choice randomization \citep{eethompson33, eewatkins89, eebridle90} \added[id=siyu]{, also known as random exploration}. In this strategy, \deleted[id=siyu]{the decision process between exploration and exploitation is corrupted by `decision noise', meaning that }high value `exploit' options are not always chosen and exploratory choices are sometimes made by chance. \added[id=siyu]{In modeling terms, random exploration works by adding 'decision noise' to the value of the options such that sub-optimal exploratory options can sometimes have a higher total score (i.e., value + noise) than the exploit option and get chosen.} \replaced[id=siyu]{Such}{In theory, such} random exploration, is surprisingly effective and, if implemented correctly, can come close to optimal performance \added[id=siyu]{in theory}\citep{eethompson33, eebridle90, eeAgrawal11, eeChapelle11}.

	It has recently been shown that humans appear to use random exploration and can increase such decision noise when it is more beneficial to explore \citep{Gershman2018, wilson2014}. In one of these tasks, known as the Horizon Task, the key manipulation is the horizon condition, i.e. the number of decisions remaining for the participant to make. Increasing the horizon makes exploration more valuable as there is more time to use the information gained by exploration to maximize future rewards. For example, if you are leaving town tomorrow (short horizon), you will probably exploit the restaurant you know and love, but if you are in town for a while (long horizon), you would be more likely to explore the new restaurant. Using such a horizon manipulation it has been shown that people's behavior is more variable in long horizons than short horizons, suggesting that they use adaptive decision noise to solve the explore-exploit dilemma \citep{wilson2014}.

	\replaced[id=siyu]{One limitation, however, is that it is difficult to tell whether what we have called 'decision noise' in previous research is truly a random noise.}{One limitation of this previous work, is that it is difficult to tell whether what is measured as decision noise is truly random.} That is, whether behavioral variability is due to intrinsic stochastic processes in the brain or whether it is due to deterministic processes that we \replaced[id=siyu]{failed to observe}{simply have not observed}. \added[id=siyu]{Decision noise as defined in previous researches are more or less a quantification of what's not predictable by the model. A missing deterministic component from the model could give rise to variability in behavior that might appear to be a random noise.} For example, in the restaurant example, my usual preference for one restaurant or another may be overruled if I see an ex romantic partner going into one of them. Avoiding an ex is a deterministic process, but if we fail to take ex's presence into account as scientists modeling the decision, then over a series of such decisions where the ex is present or not, we would mistakenly attribute the ensuing `variability' in choice to randomness.

@@

	In this paper, we investigate the extent to which the apparent randomness in random exploration can be explained by such \added[id=siyu]{unobserved} deterministic processing of the stimulus \added[id=siyu]{(will refer to as deterministic noise)}. \replaced[id=siyu]{To distinguish between stimulus-driven `deterministic noise' and non-stimulus-driven `random noise'}{In particular}, we modify the Horizon Task \citep{wilson2014} to have people face the exact same explore-exploit choice twice. If the decision is a purely deterministic function of the stimulus \added[id=siyu]{(i.e., decision noise is purely deterministic noise)}, then people's choices should be identical both times. That is, their choices should be consistent, since the stimulus is the same both times. Conversely, the more their decision is driven by \replaced[id=siyu]{random noise}{other processes, including both stochastic and unobserved deterministic processes}, the less consistent their behavior should be. \added[id=siyu]{We built a computational model to quantify the relative magnitudes of stimulus-driven deterministic noise and non-stimulus-driven random noise in driving random exploration. Our model could detect the existence of previously unobserved deterministic processes without making a priori assumptions about the deterministic process that we want to observe.} By analyzing behavior on this task in both a model-free and model-based manner, we show that at least some of the `randomness' in random exploration must come from deterministic processing of the stimulus. \replaced[id=siyu]{Our work}{This does not prove that random exploration is entirely deterministic or stochastic, but} provides a lower bound on how much deterministic processes contribute to random exploration\added[id=siyu]{, and an upper bound on how much true random noises contribute to random exploration}. \deleted[id=siyu]{This in turn sheds light on how random exploration may be implemented by amplifying the effects of task-irrelevant stimuli on choice.}

@@

	\subsection*{The Repeated-Games Horizon Task}
	We used a modified version of the `Horizon Task' \citep{wilson2014} to show the influence of stimulus-driven `deterministic noise' vs non-stimulus-driven `random noise' \replaced[id=siyu]{in explore-exploit}{on people's} decisions (Figure \ref{fig:taskfig}). In this task, participants make repeated choices between two slot machines, or `one-armed bandits,' that pay out probabilistic rewards. Because they are initially unsure as to the mean payoff of each bandit, this task requires that participants carefully balance exploration of the lesser known bandit with exploitation of the better known bandit to maximize their overall rewards.

@@

	Crucially, before people make their first choice in the Horizon Task, they are given information about the mean payoff from each bandit in the form of four example plays distributed either unequally between bandits (i.e. 1 play of one bandit, 3 plays of the other, the [1 3] condition) or equally (2 plays each, the [2 2] condition). These example plays allow us to manipulate exactly what people know about each option before they make their first choice.

	Relative to the original Horizon Task, the key modification here is to give people `repeated games,' in which they see exact same set of example plays twice in two separate games (separated by several minutes in time so as to avoid detection). By repeating the instructed plays for each game twice, we can set up a situation where (unbeknownst to the participants) they are faced with the exact same explore-exploit choice, with the exact same stimuli twice. Thus, if their behavior is a deterministic function of the stimuli, they will behave identically on the two games, that is their \replaced[id=siyu]{choices}{behavior} on the two versions of each game will be consistent. Conversely, if their behavior is not driven by a deterministic function of the stimulus, then their choices on the repeated games \replaced[id=siyu]{could}{should} be inconsistent. \replaced[id=siyu]{The extent to which participants' choices are consistent on the repeated versions of the games allow}{On average participants played 67.31 repeated games each allowing} us to quantify the extent to which \added[id=siyu]{the variability in} their behavior was \replaced[id=siyu]{driven by a deterministic process vs a random noise process}{a deterministic function of the stimulus or not}.

@@

	Before discussing the results for repeated games, we first confirm that the basic behavior in this task is consistent with our previously reported results \citep{wilson2014}. As in our previous work, we find evidence for two types of exploration in the Horizon Task: Random exploration, which is the main focus of this paper, where exploration is driven by noise, and directed exploration, where exploration is driven by information (Figure \ref{fig:modelfree}, Supplementary Figure S1).

	Random exploration is quantified in a model-free way as the probability of choosing the low mean option, $p(\mbox{low mean})$ in the equal, or [2 2], condition. This value increases with horizon, consistent with the idea that behavior is more random in horizon 6 (t(64) = 6.55, p $<$ 0.001 for [1 3], t(64) = 7.99, p $<$ 0.001 for [2 2]).  Directed exploration, is measured as the probability of choosing the more informative option $p(\mbox{high info})$ in the unequal, or [1 3], condition. Again this measure increases with horizon, showing that people are more information seeking in horizon 6 (t(64) = 6.92, p $<$ 0.001).

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/RanDetNoise_modelfree.jpg}
			\caption{
				Replication of previous findings. Both  $p(\mbox{low mean})$ (A) and $p(\mbox{high info})$ (B) increase with horizon suggesting that people use both random and directed exploration in this task.  }
			\label{fig:modelfree}
		\end{center}
	\end{figure}

	\subsection*{Model-free analysis shows that random exploration may involve both random and deterministic noise}

@@

	To gain more quantitative insight into these results, we computed theoretical values for the choice inconsistency for the purely deterministic and purely random noise cases.  For purely deterministic noise this computation is simple because people should make the exact same decisions each time in repeated games, meaning that $p(\mbox{inconsistent}) = 0$ in this case. For purely random noise, the two games should be treated independently, allowing us to compute the choice inconsistency in terms of the probability of choosing the low mean option, $p(\mbox{low mean})$, as
	\begin{equation*}
		\begin{split}
			p(\mbox{consistent}) &= p(\mbox{low mean})^2 + p(\mbox{high mean})^2\\
			&= p(\mbox{low mean})^2 + (1-p(\mbox{low mean}))^2\\ 
			\mbox{hence},\quad p(\mbox{inconsistent}) &=  
			1 - p(\mbox{consistent}) = 
			2 p(\mbox{low mean})(1-p(\mbox{low mean}))
		\end{split}
	\end{equation*}
	\added[id=siyu]{Furthermore, to account for that $p(\mbox{low mean})$ is a function of reward difference $\Delta R$ between the two bandits and the information condition $I$, we estimated the conditional probability:
		$$p(\mbox{inconsistent}|\Delta R, I) = 2 p(\mbox{low mean}|\Delta R, I)(1-p(\mbox{low mean}|\Delta R, I))$$
		Then based on the likelihood that each condition ($\Delta R$ vs $I$) occurs in the task $\rho(\Delta R, I)$, we have
		$$p(\mbox{inconsistent}) = \sum_{\Delta R, I}\rho(\Delta R, I)p(\mbox{inconsistent}|\Delta R, I)$$}

	As shown in Figure \ref{fig:mf2}, people's behavior falls in between the pure deterministic noise prediction and the pure random noise prediction. Specifically, behavior is different from pure random noise prediction in the both the [1 3] condition (t(64) = 4.83, p $<$ 0.001 for horizon 1, t(64) = 3.12 p $=$ 0.003 for horizon 6) and the [2 2] condition (t(64) = 3.92, p $<$ 0.001 for horizon 1, t(64) = 3.71, p $<$ 0.001 for horizon 6). Likewise, behavior is different from pure deterministic noise prediction in both the [1 3] condition (t(64) = 13.72, p $<$ 0.001 for horizon 1, t(64) = 16.71, p $<$ 0.001 for horizon 6) and the [2 2] condition (t(64) = 9.55, p $<$ 0.001 for horizon 1, t(64) = 17.93, p $<$ 0.001 for horizon 6). \added[id=siyu]{As a negative control of our method for estimating $p(\mbox{inconsistent})$ for purely random noise, we simulated choices using a decision model that only includes random noise (for details, see the results section on model-based analysis, and methods), and found that $p(\mbox{inconsistent})$ in this simulated data is not different from our pure random noise prediction in all horizon and uncertainty conditions (p $>$ 0.05, Supplementary Figure S3).} \replaced[id=siyu]{Together, our results suggest that both random noise and deterministic noise contribute to the choice variability in random exploration.}{This suggests that at least some of the `noise' is a deterministic function of the stimulus, } Although from this analysis it is not clear whether this deterministic noise increases with horizon or not.

@@

	To more precisely quantify the contribution of deterministic noise \added[id=siyu]{and random noise}, we turned to model fitting. We modeled behavior on the first free choice of the Horizon Task using a version of the logistic choice model in \citep{wilson2014} that was modified to differentiate \replaced[id=siyu]{between components of the noise that are deterministically driven by the stimulus (`deterministic noise') and components of the noise that are not deterministically driven by the stimulus (`random noise')}{deterministic noise from random noise}. In particular, we assume that in repeated games, the value of stimulus-driven deterministic noise is frozen whereas random noise is drawn independently both times.

>>>>

% control the fontsize
\PassOptionsToPackage{textsize=scriptsize}{todonotes}
%\usepackage[final]{changes}
\usepackage{changes}
\definechangesauthor[name={Siyu Wang}, color=red]{siyu}
\definechangesauthor[name={Bob}, color=blue]{bob}

@@

\title{\replaced[id=siyu]{Separating random and deterministic sources of computational noise in explore-exploit decisions}{The nature of decision noise in random exploration}}
\author[1,\textcurrency]{Siyu Wang}
\author[1,2,3]{Robert C. Wilson}

@@

	\newpage
	\begin{abstract}
Human decision making is inherently variable. While this variability is often seen as a sign of suboptimal behavior, recent work suggests that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. \added[id=bob]{A little randomness in these `explore-exploit' decisions is remarkably effective as it can encourage us to explore options we might otherwise ignore. In line with this idea, several studies have found evidence that people increase their behavioral variability when it is valuable to explore. A key question, however, is whether this variability in so-called `random exploration' is actually random.} That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. \added[id=bob]{By modeling behavior in this task, we were able to estimate a lower bound on the amount of variability that is deterministically driven by the stimulus and an upper bound on the amount of variability that is random.} Using this approach, we find evidence that \added[id=siyu]{at least 15$\%$} of the variability in random exploration can be accounted for by deterministic processing of the stimulus. \added[id=bob]{Conversely, this suggests that up to 85$\%$ of the variability is truly `random,' although it is still possible that this variability is driven by deterministic factors not related to the stimulus.}
% This still leaves open the possibility that much of random exploration is truly `random', but narrows the window of opportunity for \added[id=siyu]{true} stochastic processes to explain this behavior. 
\added[id=bob]{Finally, our results suggest that both deterministic and random sources of variability change proportionally to each other as the value of exploration increases, suggesting that a common noise gating mechanism may be at play in random exploration.}

@@

	Imagine trying to decide where to go to dinner on a date. You can go to your favorite restaurant, the one you both really enjoy and always go to, or you can try a new restaurant that you know nothing about. Such decisions, in which we must choose between a well-known `exploit' option and a lesser known `explore' option, are known as explore-exploit decisions.  From a theoretical perspective, making optimal explore-exploit choices, i.e. choices that maximize long-term reward, is computationally intractable in most cases \citep{eegittins74, eeBasu18}. In part because of this computational complexity, there is considerable interest in how humans and animals solve the explore-exploit dilemma in practice \added[id=siyu]{\citep{Mehlhorn15, SCHULZ20197,WILSON202149}}.
%\citep{eeauer02,eegittins79,eekrebs78,eethompson33, eewatkins89, eebridle90,eemeyer95, eebanks97,eefrank09, eesteyvers09, eelee11, eepl12,eezhang13,eedaw06, eepl11, wilson2014}

	One particularly effective strategy for solving the explore-exploit dilemma is choice randomization \citep{eethompson33, eewatkins89, eebridle90} \added[id=siyu]{, also known as random exploration}. In this strategy, \deleted[id=siyu]{the decision process between exploration and exploitation is corrupted by `decision noise', meaning that }high value `exploit' options are not always chosen and exploratory choices are sometimes made by chance. \added[id=siyu]{In modeling terms, random exploration works by adding `decision noise' to the value of the options such that sub-optimal exploratory options can sometimes have a higher total score (i.e., value + noise) than the exploit option and get chosen.} \replaced[id=siyu]{Such}{In theory, such} random exploration, is surprisingly effective and, if implemented correctly, can come close to optimal performance \added[id=siyu]{in some cases}\citep{eethompson33, eebridle90, eeAgrawal11, eeChapelle11}.

	It has recently been shown that humans appear to use random exploration and can increase  decision noise when it is more beneficial to explore \citep{Gershman2018, wilson2014} \added[id=bob]{ALSO CITE FINDLING ET AL.}. In one of these tasks, known as the Horizon Task \citep{wilson2014}, the key manipulation is the horizon condition, i.e. the number of decisions remaining for the participant to make. Increasing the horizon makes exploration more valuable as there is more time to use the information gained by exploration to maximize future rewards. For example, if you are leaving town tomorrow (short horizon), you will probably exploit the restaurant you know and love, but if you are in town for a while (long horizon), you will be more likely to explore the new restaurant. Using such a horizon manipulation it has been shown that people's behavior is more variable in long horizons than short horizons, suggesting that they use adaptive decision noise to solve the explore-exploit dilemma \citep{wilson2014}.

	\replaced[id=siyu]{One limitation of this previous research, however, is that it is difficult to tell whether what we have called `decision noise' comes from a truly random process.}{One limitation of this previous work, is that it is difficult to tell whether what is measured as decision noise is truly random.} That is, whether behavioral variability is due to genuinely stochastic processes in the brain or whether it is due to deterministic processes that we \replaced[id=siyu]{failed to observe}{simply have not observed}. \added[id=bob]{Decision noise as defined in previous research essentially quantifies the extent to which behavior cannot be predicted by a computational model. A missing deterministic component from the model could give rise to variability in behavior that might appear to be random noise.} For example, in the restaurant example, my usual preference for one restaurant or another may be overruled if I see an ex romantic partner going into one of them. Avoiding an ex is a deterministic process, but if we fail to take the ex's presence into account as scientists modeling the decision, then over a series of such decisions where the ex is present or not, we would mistakenly attribute the ensuing `variability' in choice to randomness.

@@

	In this paper, we investigate the extent to which the apparent randomness in random exploration can be explained by deterministic processing of the stimulus \added[id=bob]{(which we refer to as `deterministic noise') vs other processes, including deterministic processing of unobserved stimuli as well as truly stochastic processes (which we refer to as `random noise')}. \added[id=siyu]{To distinguish between these two types of noise} we modify the Horizon Task \citep{wilson2014} to have people face the exact same explore-exploit choice twice. If the decision is a purely deterministic function of the stimulus \added[id=bob]{(i.e., decision noise is purely deterministic noise), then people's choices should be identical for both decisions, since the stimulus is the same both times. Conversely, if the decision is a purely random function of the stimulus (i.e., decision noise is purely random noise), then people's choices will be different 50$\%$ of the time, since the random noise is different each time. In between these two extremes of purely deterministic and purely random drivers of behavioral variability, the extent to which people's decisions are consistent between the two decisions can be used to estimate the amount of deterministic and random noise.} %\added[id=siyu]{We built a computational model to quantify the relative magnitudes of stimulus-driven deterministic noise and non-stimulus-driven random noise in driving random exploration. Our model could detect the existence of previously unobserved deterministic processes that arise from the stimuli without explicitly knowing what the deterministic process is.}

    \added[id=bob]{In the following, we analyze behavior on the repeated decisions version of the Horizon Task in both a model-free and model-based manner. Our model-free analysis estimates the extent to which people's behavior is consistent across repeated versions of the same decision. By measuring how this choice consistency changes as a function of horizon, this model-free analysis offers qualitative insight into the extent to which behavioral variability is driven by deterministic vs random noise. Our model-based analysis uses a computational model of the explore-exploit decision in the Horizon Task that incorporates both noise processes. By fitting this model to the behavioral data, this model-based analysis allows us to quantify the relative size of the two sources of noise and how they change in the service of exploration.}

    % to the behavioral data.

    % of behavior that incorporates the two noise processes.

    % Our model-based analysis uses a

    % Thus, by analyzing behavior on this task in both a model-free and model-based manner, we show that at least some of the `randomness' in random exploration must come from deterministic processing of the stimulus. This does not prove that random exploration is entirely deterministic or stochastic, but provides a lower bound on how much deterministic processes contribute to random exploration\added[id=siyu]{, and an upper bound on how much true random noises contribute to random exploration. Our work sheds light on how random exploration may be partially implemented by amplifying the effects of task-irrelevant stimuli on choice.} \deleted[id=siyu]{This in turn sheds light on how random exploration may be implemented by amplifying the effects of task-irrelevant stimuli on choice.}

@@

	\subsection*{The Repeated-Games Horizon Task}
	We used a modified version of the `Horizon Task' \citep{wilson2014} to show the influence of stimulus-driven `deterministic noise' vs non-stimulus-driven `random noise' \replaced[id=siyu]{in explore-exploit}{on people's} decisions (Figure \ref{fig:taskfig}). In this task, participants make a series of \added[id=bob]{BE CAREFUL USING THE WORD REPEATED. ONLY USE IT TO REFER TO THE REPEATED GAMES - HERE I REPLACED REPEATED WITH `SERIES OF'} choices between two slot machines, or `one-armed bandits,' that pay out probabilistic rewards. Because they are initially unsure as to the mean payoff of each bandit, this task requires that participants carefully balance exploration of the lesser known bandit with exploitation of the better known bandit to maximize their overall rewards.

@@

    \added[id=bob]{THERE NEEDED TO BE MORE INFORMATION ABOUT THE HORIZON TASK HERE - REMEMBER THIS COUDL BE THE FIRST TIME PEOPLE HAVE SEEN THE TASK. The Horizon Task has two key features that together allow it to quantify explore-exploit behavior. The first of these features is  the time horizon --- the number of decisions participants will make in the future. By changing this horizon from short (1 trial) to long (6 trials), the Horizon Task allows us to control the relative value of exploration and exploitation. Just like the restaurant example in the introduction, when the horizon is short, participants should be more likely to exploit the option they believe to be best, because this leads to the highest payoff in the short term. Conversely, when the horizon is long, participants should be more likely to explore at first, because this allows them to gather information to make better choices later on. By contrasting behavior between short and long horizon conditions {\it on the very first trial}, when all else is equal, the Horizon Task allows us to quantify how behavior changes when it is valuable to explore.

    The second key feature of the Horizon Task are the four `example plays' at the start of each game that allow us to control exactly what participants know about the two bandits before they make their choice. In these example plays, participants are instructed which of the bandits to play allowing us to control how much information they have about each of the options. The instructed trials are used to set up one of two information conditions and `unequal information' or [1 3] condition, in which participants play one option once and the other three times, and an `equal information' or [2 2] condition, in which participants play both options twice.}

	% Crucially, before people make their first choice in the Horizon Task, they are given information about the mean payoff from each bandit in the form of four example plays distributed either unequally between bandits (i.e. 1 play of one bandit, 3 plays of the other, the [1 3] condition) or equally (2 plays each, the [2 2] condition). These example plays allow us to manipulate exactly what people know about each option before they make their first choice.

	Relative to the original Horizon Task, the key modification in this paper is to give people `repeated games,' in which they see exact same set of example plays twice in two separate games separated by several minutes in time so as to avoid detection. By repeating the instructed plays for each game twice, we can set up a situation where (unbeknownst to the participants) they are faced with the exact same explore-exploit choice, with the exact same stimuli twice. \added[id=bob]{Thus, if their behavior is a deterministic function of the stimuli, then they will make the same decision in both games and their choices will be consistent. Conversely, if their behavior is not driven by a deterministic function of the stimulus, then their choices on the repeated games will be inconsistent some fraction of the time.} \replaced[id=siyu]{The extent to which participants' choices are consistent on the repeated versions of the games allow}{On average participants played 67.31 repeated games each allowing} us to quantify the extent to which \added[id=siyu]{the variability in} their behavior was \replaced[id=siyu]{driven by a deterministic process vs a random noise process}{a deterministic function of the stimulus or not}.

@@

    \added[id=bob]{GENERAL COMMENT - WE NEED TO WALK PEOPLE THROUGH THE BASIC RESULTS MUCH MORE SLOWLY, OTHERWISE WILL BE CONFUSING FOR PEOPLE NEW TO THE HORIZON TASK.}

	Before discussing the results for repeated games, we first confirm that the basic behavior in this task is consistent with our previously reported results using both a model-free and model-based approach \citep{wilson2014}. %As in our previous work, we find evidence for two types of exploration in the Horizon Task: random exploration, which is the main focus of this paper, where exploration is driven by noise, and directed exploration, where exploration is driven by information. 
    \added[id=bob]{In both analyses, we focus on just the first free-choice trial in each game, where the only thing that differs between the horizon conditions is the number of choices that participants will make in the future.} %(Figure \ref{fig:modelfree}, Supplementary Figure S1).
    % As in our previous work, we find evidence for two types of exploration in the Horizon Task: random exploration, which is the main focus of this paper, where exploration is driven by noise, and directed exploration, where exploration is driven by information (Figure \ref{fig:modelfree}, Supplementary Figure S1).

    \subsubsection*{Model-free analysis}
    \added[id=bob]{
    In the model-free analysis, we quantify random and directed exploration using simple choice probabilities. Random exploration is quantified as the probability of choosing the option that has the lower average payout in the example plays in the equal, or [2 2], condition, $p(\mbox{low mean})$. The idea here is that, in the equal condition, the optimal strategy is to compute the mean payout for each bandit from the example plays and then always choose the option with the highest mean. When participants do not choose the option with the higher mean, the assumption is that this is due to some kind of `decision noise,' making the probability of choosing the low mean option a measure of behavioral variability. In this view, random exploration corresponds to an increase in $p(\mbox{low mean})$ with horizon, which is exactly what we see in the data (Figure \ref{fig:modelfree}A; t(64) = 7.99, p $<$ 0.001 for [2 2]).}

    \added[id=bob]{
    Directed exploration is quantified as the probability of choosing the more informative option $p(\mbox{high info})$ in the unequal, or [1 3], condition. The more informative option is the option played once during the example plays as choosing this option gives relatively more information (doubling the number of samples from 1 to 2) than choosing the option played three times (only increasing the number of sample by a third, from 3 to 4). In this view, directed exporation corresponds to an increase in $p(\mbox{high info})$ with horizon, which is exactly what we see in the data (Figure \ref{fig:modelfree}B; t(64) = 6.92, p $<$ 0.001).}

    \subsubsection*{Model-based analysis}

    \added[id=bob]{I'M ADDING THIS SECTION, BECAUSE IT ALLOWS US TO INTRODUCE THE MODEL EARLY. THAT WAY WHEN THE MODEL COMES IN LATER WE CAN SEE HOW IT'S CHANGED}

    \added[id=bob]{Another approach to understanding behavior in the Horizon Task is to use a computational model. In this case, we model  participants' choices on the first free-choice trial by assuming they make decisions by computing the difference in value (or utility) $\Delta Q$ between the right and left options, choosing right when $\Delta Q > 0$ and left otherwise.  Specifically, we write
	\begin{equation}
    \label{eq:origmodel}
		\Delta Q= \Delta R+A \Delta I+b+n
	\end{equation}
    where, the experimentally controlled variables are $\Delta R=R_{right}-R_{left}$, the difference between the mean of rewards shown on the forced trials, and $\Delta I$, the difference in information available for playing the two options on the first free-choice trial. For simplicity, and because information is manipulated categorically in the Horizon Task, we define $\Delta I$ to be +1 if one reward is drawn from the right option and three are drawn from the left in the [1 3] condition, -1 if one from the left and three from the right, and in [2 2] condition, $\Delta I$ is 0. $n_{det}$ denotes the deterministic noise, which is identical on the repeat versions of each game; and $n$ denotes decision noise, which, in this version of the model is a combination of deterministic and random noise. $n$ is assumed to come from a logistic distributions with mean 0 and standard deviations $\sigma$. }

    \added[id=bob]{The free parameters of this model are: the information bonus $A$, which controls the level of directed exploration; the noise standard deviation, $\sigma$, which controls the level of random exploration, and the spatial bias, $b$, which determines the extent to which participants prefer the option on the right. These free parameters are fit separately for each participant in each horizon and information condition, allowing us to test whether directed and random exploration increase with horizon. Consistent with previous research, we find that this is indeed the case (Figure \ref{fig:modelfree}C, D; STATS FOR CHANGE IN INFO BONUS AND DECISION NOISE). }

    \added[id=bob]{Taken together our model-free and model-based analyses agree with previous findings showing increased behavioral variability and increased information seeking in the long horizon condition consistent with humans using random and directed exploration. However, for random exploration, this previous analysis cannot distinguish between deterministic and random sources of noise. For this we analyze the extent to which people's choices are consistent on the repeated games.}

	%  In this analysis, random exploration is quantified in a model-free way as the probability of choosing the option that has the lower average payout in the example plays in the equal, or [2 2], condition, $p(\mbox{low mean})$. The idea here is that in the equal condition, the optimal strategy is to compute the average payout for each bandit from the example plays and then always choose the option with the highest average. When participants do not choose the higher average option, the assumption is that this is due to some kind of `decision noise.' In this view, random exploration corresponds to an increase in $p(\mbox{low mean})$ with horizon, which is exactly what we see in the data (Figure \ref{fig:modelfree}A; t(64) = 6.55, p $<$ 0.001 for [1 3], t(64) = 7.99, p $<$ 0.001 for [2 2]).}

 % Directed exploration, is measured as the probability of choosing the more informative option $p(\mbox{high info})$ in the unequal, or [1 3], condition. Again this measure increases with horizon, showing that people are more information seeking in horizon 6 than horizon 1 (t(64) = 6.92, p $<$ 0.001).  \added[id=bob]{Together, these results are consistent with the idea that people use both random exploration and directed exploration and increase both behavioral variability and information seeking when the time horizon is long.}

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/RanDetNoise_modelfree.jpg}
			\caption{\added[id=bob]{ADD C AND D PANELS FOR MODEL-BASED RESULTS HERE}
				Replication of previous findings. Both  $p(\mbox{low mean})$ (A) and $p(\mbox{high info})$ (B) increase with horizon suggesting that people use both random and directed exploration in this task.  }
			\label{fig:modelfree}
		\end{center}
	\end{figure}

	\subsection*{Model-free analysis of repeated games suggests that random exploration involves both random and deterministic noise}

@@

	To gain more quantitative insight into these results, we computed theoretical values for the choice inconsistency for the purely deterministic and purely random noise cases.  For purely deterministic noise this computation is simple because people should make the exact same decisions each time in repeated games, meaning that $p(\mbox{inconsistent}) = 0$ in this case. For purely random noise, the two games should be treated independently, allowing us to compute the choice inconsistency in terms of the probability of choosing the low mean option, $p(\mbox{low mean})$, as
	\begin{equation*}
		\begin{split}
			p(\mbox{consistent}) &= p(\mbox{low mean})^2 + p(\mbox{high mean})^2\\
			&= p(\mbox{low mean})^2 + (1-p(\mbox{low mean}))^2\\ 
			\mbox{hence},\quad p(\mbox{inconsistent}) &=  
			1 - p(\mbox{consistent}) = 
			2 p(\mbox{low mean})(1-p(\mbox{low mean}))
		\end{split}
	\end{equation*}
	\added[id=siyu]{Furthermore, to account for the fact that $p(\mbox{low mean})$ is a function of reward difference $\Delta R$ between the two bandits and the information condition $I$, we estimated the conditional probability:
		$$p(\mbox{inconsistent}|\Delta R, I) = 2 p(\mbox{low mean}|\Delta R, I)(1-p(\mbox{low mean}|\Delta R, I))$$
	Then based on the likelihood that each condition ($\Delta R$ vs $I$) occurs in the task $\rho(\Delta R, I)$, we have
		$$p(\mbox{inconsistent}) = \sum_{\Delta R, I}\rho(\Delta R, I)p(\mbox{inconsistent}|\Delta R, I)$$}

	As shown in Figure \ref{fig:mf2}, people's behavior falls in between the pure deterministic noise prediction and the pure random noise prediction. Specifically, behavior is different from the pure random noise prediction in the both the [1 3] condition (t(64) = 4.83, p $<$ 0.001 for horizon 1, t(64) = 3.12 p $=$ 0.003 for horizon 6) and the [2 2] condition (t(64) = 3.92, p $<$ 0.001 for horizon 1, t(64) = 3.71, p $<$ 0.001 for horizon 6). Likewise, behavior is different from pure deterministic noise prediction in both the [1 3] condition (t(64) = 13.72, p $<$ 0.001 for horizon 1, t(64) = 16.71, p $<$ 0.001 for horizon 6) and the [2 2] condition (t(64) = 9.55, p $<$ 0.001 for horizon 1, t(64) = 17.93, p $<$ 0.001 for horizon 6). \added[id=bob]{As a negative control of our method for estimating $p(\mbox{inconsistent})$ for purely random noise, we simulated choices using a decision model that only includes random noise (Equation \ref{eq:origmodel}), and found that $p(\mbox{inconsistent})$ in this simulated data is not different from our pure random noise prediction in all horizon and uncertainty conditions (p $>$ 0.05, Supplementary Figure S3).} \replaced[id=siyu]{Together, our results suggest that both random noise and deterministic noise contribute to the choice variability in random exploration.}{This suggests that at least some of the `noise' is a deterministic function of the stimulus, }. However,  how each of these types of noise change with horizon is difficult to discern.

@@

	To more precisely quantify the contribution of deterministic noise \added[id=siyu]{and random noise}, we turned to model fitting. We modeled behavior on the first free choice of the Horizon Task using a version of the logistic choice model in \citep{wilson2014} (Equation \ref{eq:origmodel}) that was modified to differentiate \replaced[id=siyu]{between components of the noise that are deterministically driven by the stimulus (`deterministic noise') and components of the noise that are not deterministically driven by the stimulus (`random noise')}{deterministic noise from random noise}. In particular, we assume that in repeated games, the value of stimulus-driven deterministic noise is frozen whereas random noise is drawn independently both times.
